---
title: "STAT 415 Project"
author: "Bryan Veres, Casey Ng"
date: "3/10/2021"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 4
    toc_float: true
    smooth_scroll: false
toc-title: "Table of Contents"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# 1. Research question and context
Basketball is a relatively simple game, you shoot a ball into a hoop and you get points. If you end up with more points than the other team, you win. At the highest level of play, a basketball game is 48 minutes long. In a game, there’ll be about 15 players that you can switch around throughout the game. Obviously you’ll want your better players on the court longer to increase your chances of winning but not in the whole game which will exhaust them. Another important aspect in scoring points is shooting percentage. For example, scoring 20 points on 10 shots is more impactful than scoring 20 points on 20 shots. We wanted to research if we can use minutes per game played and true shooting percentage to predict the number of points made per 36 minutes.

# 2. Data set
https://github.com/fivethirtyeight/nba-player-advanced-metrics?fbclid=IwAR0N-fvRFmpBKqHvWP7_bTdgPYtxp0jWeX1ssV4SXfoAK8MJ1n9kIWNzd1I

# 3. Discuss data and how it was collected
The data was taken from 538, a website that uses statistical analysis to tell stories about politics, science, sports, etc. NBA data is usually collected using cameras that record the movement of the ball and every player on the court. NBA advanced stats can also usually be found on their website or other websites that archive basketball stats like basketball reference. For the variables we are using, minutes per game is pretty simple, the average number of minutes a player is on the court per game for a season. True shooting percentage is taken from the number of points made and divided by 2 times by true shooting attempts (field goal attempts +0.44*free throw attempts). We are using this statistic since it's the most accurate measure of shooting efficiency that takes field goals, three point field goals and free throws into account. Points per 36 minutes takes point average, divided by minutes played per game and then multiplied by 36. Because this data set is extremely large we decided to filter the data to just 2019, the last complete 82 game regular season.



```{r, include=FALSE}
library(tidyverse)
library(rjags)
library(bayesplot)
library(bayesboot)
```

```{r}
data <- read_csv("C:/Users/Casey/Desktop/STAT415/nba-data-historical.csv")
data <- read_csv("C:/Users/Bryan Veres/Desktop/nba-data-historical.csv")
```

```{r}
data <- data %>% 
  filter(year_id == 2019 & !is.na(`TS%`))
```

# 4.Propose a Bayesian model
Given that we wanted to see if minutes per game and true shooting percentage are good predictors of points per 36 minutes, which are all quantitative variables, we propose that a Bayesian analysis of multiple regression would best model this relationship.

# 5. Discuss choice of prior and likelihood
After looking at plots of the raw data for each variable, we could say that the distributions of the data follows a roughly normal distribution. But, with relatively limited information on the true distributions of the data, we decided to choose relatively non-informative priors, assigning normal distributions with means of 0 and large prior SD’s of about 20 to each $\beta$ value in the regression line equation. For the value of $\sigma$, we assigned a Gamma prior with parameters $\alpha$ = $\beta$ = 0.001. Given an equation for a regression line, representing the predicted average value of y, the points per 36 minutes, and a value of $\sigma$, our likelihood is represented by a Normal($\beta_0$ + $\beta_1$*$x_{min}$ + $\beta_2$($x_{ts}$-mean($x_{ts}$)), $\sigma$) distribution, where $\beta_0$ represents the intercept of the regression line (when the minutes per game and true shooting percentage are both 0), $\beta_1$ represents the average change in points per 36 minutes for each additional minute per game after adjusting for true shooting percentage, $\beta_2$ represents the average change in points per 36 minutes for each additional percentage point in true shooting percentage after adjusting for minutes per game, and $x_{min}$ and $x_{ts}$ represent the values of the observed minutes per game and true shooting percentage, respectively, for a given player. This likelihood function follows as the assumptions of multiple linear regression models assume that the values of the response follow a Normal distribution.

As a note, we first tried the model without centering the true shooting percentage. However, this model did not work well with the Gibbs sampling as the diagnostic plots were not good for values of $\beta_0$ and $\beta_1$, as the autocorrelations decayed very slowly. Once the true shooting percentage was centered, the autocorrelations for both Beta0 and Beta1 decayed fairly fast, along with $\beta_2$.

# 6. Fitting the Model
```{r}
modelString <- "model {

  ## sampling
  for (i in 1:N) {
    y[i] ~ dnorm(beta0 + beta1*x_min[i] + beta2*(x_ts[i]-mean(x_ts)), sigma2)
  }
  
  ## Priors
  beta0 ~ dnorm(mu0, g0)
  beta1 ~ dnorm(mu1, g1)
  beta2 ~ dnorm(mu2, g2)
  sigma2 ~ dgamma(a, b)
  sigma <- sqrt(pow(sigma2, -1))

}"
```

```{r}
y <- as.vector(data$`P/36`)
x_min <- as.vector(data$MPG)
x_ts <- as.vector(data$`TS%`)
N <- length(y)
the_data <- list("y" = y,
                 "x_min" = x_min, "x_ts" = x_ts,
                 "N" = N,
                 "mu0" = 0, "g0" = 0.0025,
                 "mu1" = 0, "g1" = 0.0025,
                 "mu2" = 0, "g2" = 0.0025,
                 "a" = 0.001, "b" = 0.001)
```

```{r}
model <- jags.model(textConnection(modelString),
                    data = the_data,
                    n.chains = 5)
```

```{r}
update(model, 1000, progress.bar = "none");

posterior_sample <- coda.samples(model,
                                 variable.names = c("beta0", "beta1", "beta2", "sigma2"),
                                 n.iter = 10000, progress.bar="none")
summary(posterior_sample)
```

# 7. Posterior inference
The posterior distribution of Beta1, the slope for minutes per game in the population regression line, is approximately Normal with a posterior mean of 0.15623 and a posterior SD of 0.021182. There is a 95% posterior probability that the slope for minutes per game in the population regression line is between 0.1147 and 0.1977. That is, there is a posterior probability of 95% that, after adjusting for true shooting percentage, an increase in 1 minute per game is associated with between a 0.1147 and 0.1977 increase in points per 36 minutes, on average.

The posterior distribution of Beta2, the slope for true shooting percentage in the population regression line, is approximately Normal with a posterior mean of 0.23819 and a posterior SD of 0.016365. There is a 95% posterior probability that the slope for true shooting percentage in the population regression line is between 0.2062 and 0.27023. That is, there is a posterior probability of 95% that, after adjusting for minutes per game, an increase in 1 percentage point of true shooting percentage is associated with between a 0.2062 and 0.27023 increase in points per 36 minutes, on average.


```{r}
plot(posterior_sample)
```

```{r}
params <- as.matrix(posterior_sample)
beta0 <- params[,1]
beta1 <- params[,2]
beta2 <- params[,3]
sigma2 <- params[,4]
```

```{r}
mcmc_acf(posterior_sample)
```

```{r}
mcmc_trace(posterior_sample)
```

```{r}
plotPost(beta0)
plotPost(beta1)
plotPost(beta2)
plotPost(sigma2)
```

# 8. Posterior prediction
For posterior prediction, we found posterior predictive distributions for current players using their current MPG’s and TS% from this year, namely Stephen Curry and Draymond Green.

Using Steph Curry’s current MPG and TS%, we found that there is a posterior predictive probability of 95% that the P/36 of a player who plays 34.1 minutes per game and has a true shooting percentage of 64.6 is between 19.11 and 20.57. Roughly, 95% of players with 34.1 MPG and a 64.6 TS% score between 19.11 and 20.57 P/36.

Using Draymond Green’s’ current MPG and TS%, we found that there is a posterior predictive probability of 95% that the P/36 of a player who plays 29.6 minutes per game and has a true shooting percentage of 46.4 is between 14.16 and 15.45. Roughly, 95% of players with 29.6 MPG and a 46.4 TS% score between 14.16 and 15.45 P/36.

## Stephen Curry
```{r}
min_new = 34.1
ts_new = 64.6
ynew = beta0 + beta1*min_new + beta2*(ts_new - mean(x_ts)) + sigma2*rnorm(50000, 0, 1)
hist(ynew, freq = FALSE, main=paste("Mean=", mean(ynew), ", SD=", sd(ynew)))
lines(density(ynew))

quantile(ynew, c(0.025, 0.975))
```

## Draymond Green
```{r}
min_new = 29.6
ts_new = 46.4
ynew = beta0 + beta1*min_new + beta2*(ts_new - mean(x_ts)) + sigma2*rnorm(50000, 0, 1)
hist(ynew, freq = FALSE, main=paste("Mean=", mean(ynew), ", SD=", sd(ynew)))
lines(density(ynew))

quantile(ynew, c(0.025, 0.975))
```

# 9. Posterior predictive checking or model comparison
Firstly, we perform posterior predictive checking by computing posterior predictive intervals for P/36 for a variety of values of minutes per game, adjusting for the true shooting percentage, using the sample mean TS%, sample 25th percentile and sample 75th percentile as the values of x_ts. Based on the scatterplots we see that most observed values of P/36 fall near/along the centers of the predictive intervals. However, since the predictive intervals have very small standard deviations, likely due to the large observed sample size, most values don’t fall within the intervals.

Next, performing posterior predictive checking similarly for the true shooting percentage by computing posterior predictive intervals for P/36 for a variety of values of true shooting percentage, adjusting for minutes per game, by using the sample mean MPG, sample 25th percentile and sample 75th percentile as the values of x_min. Based on the scatterplots, similarly to the plots for minutes per game, we see that most observed values of P/36 fall around the predictive intervals, but not within the intervals due to the intervals having such small standard deviations. 

Although, the predicted intervals seem to follow the overall trend of the data, we model might require adjustment given that most of the sample data falls outside of the intervals.

## Minutes Per Game
### TS% = Sample Mean TS%
```{r}
plot(data$MPG, data$`P/36`, xlab = "Minutes Per Game", ylab = "Points per 36 Minutes",
     main = "95% Posterior Prediction Intervals")

xnew = seq(0, 37.5, 2.5)

for (xx in xnew) {
  ynew = beta0 + beta1 * (xx) + sigma2 * rnorm(50000, 0, 1)
  segments(x0 = xx, y0 = quantile(ynew, 0.025), x1 = xx,
           y1 = quantile(ynew, 0.975), col = "orange", lwd=2)
  points(x = xx, y = mean(ynew), pch = 19, col = "orange")
}
```

### TS% = Sample 25th Percentile TS%
```{r}
plot(data$MPG, data$`P/36`, xlab = "Minutes Per Game", ylab = "Points per 36 Minutes",
     main = "95% Posterior Prediction Intervals")

xnew = seq(0, 37.5, 2.5)

for (xx in xnew) {
  ynew = beta0 + beta1 * (xx) + beta2 * (quantile(x_ts, 0.25) - mean(x_ts)) + sigma2 * rnorm(50000, 0, 1)
  segments(x0 = xx, y0 = quantile(ynew, 0.025), x1 = xx,
           y1 = quantile(ynew, 0.975), col = "orange", lwd=2)
  points(x = xx, y = mean(ynew), pch = 19, col = "orange")
}
```

### TS% = Sample 75th Percentile TS%
```{r}
plot(data$MPG, data$`P/36`, xlab = "Minutes Per Game", ylab = "Points per 36 Minutes",
     main = "95% Posterior Prediction Intervals")

xnew = seq(0, 37.5, 2.5)

for (xx in xnew) {
  ynew = beta0 + beta1 * (xx) + beta2 * (quantile(x_ts, 0.75) - mean(x_ts)) + sigma2 * rnorm(50000, 0, 1)
  segments(x0 = xx, y0 = quantile(ynew, 0.025), x1 = xx,
           y1 = quantile(ynew, 0.975), col = "orange", lwd=2)
  points(x = xx, y = mean(ynew), pch = 19, col = "orange")
}
```


## True Shooting Percentage
### Minutes Per Game = Sample Mean MPG
```{r}
plot(data$`TS%`, data$`P/36`, xlab = "True Shooting Percentage", ylab = "Points per 36 Minutes",
     main = "95% Posterior Prediction Intervals", xlim = c(0, 100))

xnew = seq(0, 100, 5)

for (xx in xnew) {
  ynew = beta0 + beta1*mean(x_min) + beta2 * (xx - mean(x_ts)) + sigma2 * rnorm(50000, 0, 1)
  segments(x0 = xx, y0 = quantile(ynew, 0.025), x1 = xx,
           y1 = quantile(ynew, 0.975), col = "orange", lwd=2)
  points(x = xx, y = mean(ynew), pch = 19, col = "orange")
}
```

### Minutes Per Game = Sample 25th Percentile MPG
```{r}
plot(data$`TS%`, data$`P/36`, xlab = "True Shooting Percentage", ylab = "Points per 36 Minutes",
     main = "95% Posterior Prediction Intervals", xlim = c(0, 100))

xnew = seq(0, 100, 5)

for (xx in xnew) {
  ynew = beta0 + beta1*quantile(x_min, 0.25) + beta2 * (xx - mean(x_ts)) + sigma2 * rnorm(50000, 0, 1)
  segments(x0 = xx, y0 = quantile(ynew, 0.025), x1 = xx,
           y1 = quantile(ynew, 0.975), col = "orange", lwd=2)
  points(x = xx, y = mean(ynew), pch = 19, col = "orange")
}
```

### Minutes Per Game = Sample 75th Percentile MPG
```{r}
plot(data$`TS%`, data$`P/36`, xlab = "True Shooting Percentage", ylab = "Points per 36 Minutes",
     main = "95% Posterior Prediction Intervals", xlim = c(0, 100))

xnew = seq(0, 100, 5)

for (xx in xnew) {
  ynew = beta0 + beta1*mean(x_min) + beta2 * (xx - mean(x_ts)) + sigma2 * rnorm(50000, 0, 1)
  segments(x0 = xx, y0 = quantile(ynew, 0.025), x1 = xx,
           y1 = quantile(ynew, 0.975), col = "orange", lwd=2)
  points(x = xx, y = mean(ynew), pch = 19, col = "orange")
}
```


# 10. Sensitivity Analysis
Because the posterior distribution still has really small standard deviations that makes it hard to fit the data in the intervals, we want to see if there would be any significant change if we increase the SD in our prior.We specifically want to see if there is any significant difference in the posterior SD when we change the prior SD $\beta_0$, $\beta_1$, and $\beta_2$ to 100.

```{r,echo=FALSE}
modelString <- "model {

  ## sampling
  for (i in 1:N) {
    y[i] ~ dnorm(beta0 + beta1*x_min[i] + beta2*(x_ts[i]-mean(x_ts)), sigma2)
  }
  
  ## Priors
  beta0 ~ dnorm(mu0, g0)
  beta1 ~ dnorm(mu1, g1)
  beta2 ~ dnorm(mu2, g2)
  sigma2 ~ dgamma(a, b)
  sigma <- sqrt(pow(sigma2, -1))

}"
```

```{r,echo=FALSE}
y <- as.vector(data$`P/36`)
x_min <- as.vector(data$MPG)
x_ts <- as.vector(data$`TS%`)
N <- length(y)
the_data <- list("y" = y,
                 "x_min" = x_min, "x_ts" = x_ts,
                 "N" = N,
                 "mu0" = 0, "g0" = 0.0001,
                 "mu1" = 0, "g1" = 0.0001,
                 "mu2" = 0, "g2" = 0.0001,
                 "a" = 0.001, "b" = 0.001)
```

```{r,echo=FALSE}
model <- jags.model(textConnection(modelString),
                    data = the_data,
                    n.chains = 5)
```

```{r,echo=FALSE}
update(model, 1000, progress.bar = "none");

posterior_sample <- coda.samples(model,
                                 variable.names = c("beta0", "beta1", "beta2", "sigma2"),
                                 n.iter = 10000, progress.bar="none")
summary(posterior_sample)
```

When comparing the summary of the posterior sample to our original model, the results are approximately the same. This shows that the posterior sample doesn't seem to make any major changes when adjusting the prior.

# 11. Conclusions
From looking at our posterior predictive intervals, while the intervals for both MPG and TS% seem to fit with the shape of the data, most of the data values don't seem to fall within both of our posterior predictive intervals. While it's plausible that MPG and TS% can be used to predict P/36, but the posterior predictive interval seems to be too small to fit most of the data. One of the biggest explanations for why the interval is too small is because since there's alot of data points, the standard deviation is bound to be small. Other reasons may include is that star players who may score more will play more minutes, a lot of starting caliber players in the NBA will also generally have a high MPG. Another reason may include that some players that get many minutes may not have scoring as their priority. One of the best examples is Draymond Green, a more pass first type player due to not being as great of a scorer. As a result, his P/36 is only 8.4 despite averaging 31.3 minutes a game. Another explanation for the bad fit may be because TS% seems to be very similar between most stars and regular players. And since more regular players are going to have less opportunities to shoot compared to stars, they will have a lower P/36 as a result. Looking at our data, there also seem to be a lot of outliers that possibly affect the model. A good example of this is Rockets legend Zhou Qi, who played one minute the whole season and scored 2 points on 1 shot, giving him a P/36 of 73.6 and a TS% of 100%. In conclusion, minutes per game and true shooting percentage are not the best predictors for points per 36 minutes.